Extended Data

### 2021-04-27 doc/manuscript/Dow_Extended_Data.Rmd
implement tentative Extended Data items
#52

# EXTENDED DATA 
##tables
table_species="Extended Data Table 1"  
sample_size_by_year= "Extended Data Table 2"
mean_growth_parameters_table="Extended Data Table 3" 
chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions

##figures
growth_curves_all = "Extended Data Figure 1"  
climwin = "Extended Data Figure 2"
tree_rings_TMIN = "Extended Data Figure 3"

## table captions
species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
sample_size_by_year_caption <- "Extended Data Table 2 | Sample size by year "
mean_growth_parameters_caption = "Extended Data Table 3 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "

## figure captions
growth_curves_all_caption  <- " Extended Data Figure 1 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth. Each line represents one year's growth for a given tree, fit with McMahon model. "

climwin_caption <- "Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest. Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF). Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb. 19- May 21 at SCBI, March 19 - May 07 at HF)."

tree_rings_TMIN_short_caption <- "Extended Data Figure 3 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America. Chronologies are grouped by xylem porosity and ordered by mean April T_max. Chronology details are given in the Supplementary Information. (**NOTE: Figure still needs some work. Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"

@@

*This figure needs some work if we choose to include it. It would also probably make sense to include years for which leaf phenology is available.*

@@

@@

![`r tree_rings_TMIN_short_caption`](tables_figures/tree_rings_tmin.png){ width=90% }

>>>>

# EXTENDED DATA 
##tables
EDT_species="Extended Data Table 1"  
EDT_mean_growth_parameters="Extended Data Table 2" 
EDT_tree_ring_summary="Extended Data Table 3"

sample_size_by_year= "XX"
#chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions

##figures
EDF_map = "Extended Data Figure 1" 
EDF_growth_curves_all = "Extended Data Figure 2"  
EDF_climwin = "Extended Data Figure 3"
EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
EDF_tree_rings_TMIN = "Extended Data Figure 7"

## table captions
species_table_caption <-"Extended Data Table 1 | Dominant ring- and diffuse-porous species at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest, along with sample sizes included in this analysis."
sample_size_by_year_caption <- "TO BE REMOVED | Sample size by year "
mean_growth_parameters_caption = "Extended Data Table 2 | Summary of parameters describing the phenology and rate of growth for ring- and diffuse- porous species at SCBI and Harvard Forest. "
tree_ring_summary_caption =  "Extended Data Table 3 | Summary of tree-ring data and analysis ( including species, n chronologies (= n sites), n positively correlated with April T, n sig.)"

## figure captions
growth_curves_all_caption  <- " Extended Data Figure 2 | Growth trajectories for ring- and diffuse-porous trees, as both relative and cumulative fractions of total annual growth. Each line represents one year's growth for a given tree, fit with McMahon model. " #needs wordsmithing!!!

climwin_caption <- "Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at SCBI and Harvard Forest. Shown are matrices of linear coefficients of first-order linear regressions between temperature and $DOY_{25}$, where Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (ring-porous: May 5 at SCBI, May 13 at HF; diffuse-porous: June 4 at SCBI, June 14 at HF). Black circles indicate the critical $T_{max}$ window (ring-porous:  March 22- April 9 at SCBI, April 2 - May 07 at HF; diffuse-porous: Feb. 19- May 21 at SCBI, March 19 - May 07 at HF)." #info subject to change!!!

pheno_Tmax_climwin_caption <- "**Extended Data Figure 4 | Relationship between growth parameters and mean maximum temperature within the critical temperature window (CTW).** CTW was defined using *climwin*. For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmax_AprilMay_caption <- "**Extended Data Figure 5 | Relationship between growth parameters and mean maximum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!

pheno_Tmin_AprilMay_caption <- "**Extended Data Figure 6 | Relationship between growth parameters and mean minimum temperature in April (SCBI) or May (HF).** For each observed climwin mean temperature value, the posterior mean of the fitted day of year is represented by the solid blue line and 95% credible intervals are represented by bands.(**NEEDS MORE DETAIL**)" #needs attention!!!

tree_rings_TMIN_caption <- "Extended Data Figure 7 | Sensitivity of annual growth, as derived from tree-rings, to monthly minimum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1). Chronologies are grouped by xylem porosity and ordered by mean April T_max. Chronology details are given in the Supplementary Information. (**NOTE: Figure still needs some work. Chronology numbers are off.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!! #info subject to change!!!

@@

*This figure PROBABLY NEEDS TO GO. It needs some work if we choose to include it. *

@@

## (PLACEHOLDER FOR `r tree_ring_summary_caption` )

## (PLACEHOLDER FOR MAP OF STUDY SITES -- [GITHUB ISSUE #54](https://github.com/EcoClimLab/growth_phenology/issues/54))

@@

![`r pheno_Tmax_climwin_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

![`r pheno_Tmax_AprilMay_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may.png){ width=90% }

![`r pheno_Tmin_AprilMay_caption`](tables_figures/pheno_Tsensitivity_combo_patchwork_april_may_tmin.png){ width=90% }

![`r tree_rings_TMIN_caption`](tables_figures/tree_rings_tmin.png){ width=90% }

### 2021-04-27 doc/manuscript/Dow_manuscript_Nature.Rmd
implement tentative Extended Data items
#52

pheno_Tsensitivity_caption <- "Figure 2 | Relationship between growth parameters and climwin mean temperatures. For each 
observed climwin mean temperature value, the posterior mean of the fitted day of year
is represented by the solid blue line and 95% credible intervals are represented by bands. (**SHOULD WE SHOW ONLY THE DOY TRENDS and Lpgs (top 2 rows), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)"

DOYtiming_caption <- "Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available). Panels (a-b) show leaf phenology... Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site. Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50). )"

tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America. Chronologies are grouped by xylem porosity and ordered by mean April T_max. Chronology details are given in the Supplementary Information. (NOTE: Figure still needs some work. **COLOR SCALE IS  A BIT OFF.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)"

# EXTENDED DATA  (this chunk must be repeated in Dow_Extended_Data.Rmd)
##tables
table_species="Extended Data Table 1"  
sample_size_by_year= "Extended Data Table 2"
mean_growth_parameters_table="Extended Data Table 3" 
chronologies_table="S4" #!!! Hand-numbered in tree-ring figure captions

##figures
growth_curves_all = "Extended Data Figure 1"  
climwin = "Extended Data Figure 2"
tree_rings_TMIN = "Extended Data Figure 3"

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
However, little is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood, which has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with *previous* observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] must be allocated primarily to relatively short-lived C pools and therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (`r growth_curves_all`, `r mean_growth_parameters_table`).**
Across all trees and years, the mean $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively). 
Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species.

@@

For the majority of weekly time windows from Jan. 1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r climwin`). 
This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
The strongest relationships, including but not limited to the optimal time windows (`r mean_growth_parameters_table`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.

@@

**To understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies for `r n_chronologies_SCBI` species at SCBI, `r n_chronologies_HF` species at Harvard Forest.**
*(The following should be condensed:)*
Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig. `r tree_rings`, `r tree_rings_TMIN`).
Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig. `r tree_rings`).
In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
In the prior months, the generally negative correlations to $T_{max}$ were not significant, excepting significant correlations of one species at each site  (ring-porous *Carya tomentosa* in March at SCBI; diffuse-porous *Acer rubrum* in April at Harvard Forest).
Starting in May at SCBI, significant correlations to $T_{max}$ were common, occurring for nine of the eleven species in May and 3-7 species for each month June through August.
At Harvard Forest, there was only one statistically significant negative correlation to post-May $T_{max}$ (*Quercus rubra* in June).

@@

Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r tree_rings_TMIN`). 
There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.

@@

Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r table_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census (`r sample_size_by_year`) *(tags_per_year_preclean in data folder)*. 
Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (`r table_species`).

@@

### Tree-rings
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (`r table_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r table_species`).

@@

All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time [@cook_calculating_1997]. Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology [@jones_estimating_1997]. The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c. the inner 30–50 yr of each record depending on full chronology length) dropped below three trees. Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices [@cook_time_1985]. We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].* 
We defined chronology start year (`r table_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.

@@

The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window. 
Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r mean_growth_parameters_table`).
*The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic_summary`), because we were interested in the influence of temperature on the initiation of woody growth. 
We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019]. Determining the climate variables that most strongly control all parameters is beyond the scope of this analysis.

>>>>

pheno_Tsensitivity_caption <- "Figure 2 | Relationship between growth parameters and climwin mean temperatures. For each 
observed climwin mean temperature value, the posterior mean of the fitted day of year
is represented by the solid blue line and 95% credible intervals are represented by bands. (**SHOULD WE SHOW ONLY THE DOY TRENDS and Lpgs (top 2 rows), GIVEN LACK OF SIGNIFICANT TRENDS IN THE OTHERS AND BECAUSE WE NEED TO KEEP FIGURES COMPACT?**)" #needs attention!!!

DOYtiming_caption <- "Figure 3 | Leaf (a,b) and stem growth (c,d) phenology at the Smithsonian Conservation Biology Institute (a,c) and Harvard Forest (b,d) from 1998 to 2020 (when available). Panels (a-b) show leaf phenology... Panels (c-d) show the dates at which stem growth milestones were acheived, on average, for sampled populations of ring-porous and diffuse-porous trees at each site. Mean temperature was calculated for each wood-type/site combination over the respective critical $T_{max}$ window, then turned into a ratio and assigned a color on a gradient where the coldest year in the sample is blue and the warmest is red. (**SHOULD WE COMBINE LEAF AND STEM PHENOLOGY ON A SINGLE PLOT WITH SEPARATE Y-AXES?** See [GitHub Issue #50](https://github.com/EcoClimLab/growth_phenology/issues/50). )" #needs attention!!!

tree_rings_caption <- "Figure 4 | Sensitivity of annual growth, as derived from tree-rings, to monthly maximum temperatures, for 207 chronologies from 114 sites across eastern North America (Extended Data Figure 1). Chronologies are grouped by xylem porosity and ordered by mean April T_max. Chronology details are given in the Supplementary Information. (NOTE: Figure still needs some work. **COLOR SCALE IS  A BIT OFF.**  See [GitHub Issue #49](https://github.com/EcoClimLab/growth_phenology/issues/49).)" #needs attention!!! #info subject to change!!!

# EXTENDED DATA  (this chunk must be repeated in Dow_Extended_Data.Rmd)
##tables
EDT_species="Extended Data Table 1"  
EDT_mean_growth_parameters="Extended Data Table 2" 
EDT_tree_ring_summary="Extended Data Table 3"

##figures
EDF_map = "Extended Data Figure 1" 
EDF_growth_curves_all = "Extended Data Figure 2"  
EDF_climwin = "Extended Data Figure 3"
EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
EDF_pheno_Tmax_AprilMay = "Extended Data Figure 5"
EDF_pheno_Tmin_AprilMay = "Extended Data Figure 6"
EDF_tree_rings_TMIN = "Extended Data Figure 7"

@@

As the climate changes, warmer spring temperatures are causing earlier leaf-out[@jeong_phenology_2011] and commencement of net carbon dioxide (CO~2~) sequestration[REFS] in temperate deciduous forests, resulting in a tendency towards increased growing season length[@menzel_growing_1999;@menzel_european_2006;@ibanez_forecasting_2010;@jeong_phenology_2011] and annual CO~2~ sequestration[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016].
However, little is known about how spring temperatures affect tree stem growth, which consumes only a modest portion of the carbon fixed annually[@anderson-teixeira_carbon_2021] but sequesters that carbon in wood, which has a long residence time in the ecosystem [@xue_global_2017;@russell_residence_2014] or as wood products[@profft_forest_2009].
Here, we show that in two temperate deciduous forests, warmer maximum spring temperatures consistently shifted woody growth milestones earlier but had mixed, weak effects on peak growing season length, maximum daily growth rates, and annual growth.
The latter finding is confirmed on the centennial scale by `r n_chronologies` tree-ring chronologies from `r n_cores_sites` forests across eastern North America, which show that annual growth is by far more sensitive to temperatures during the peak growing season than in the spring.
These findings align with eddy-covariance records showing that annual CO~2~ sequestration is more strongly linked to conditions in the peak growing season than to growing season length[@fu_maximum_2019], with observations that warm springs result in earlier leaf senescence[@zohner_ongoing_2019; @zani_increased_2020], and with evidence that forest CO~2~ sequestration is sink-limited[@zani_increased_2020].
They contrast, however, with the expectation *in most models[@zani_increased_2020]* that an earlier start to growing season would significantly increase the rate of biomass accumulation and thereby enhance the CO~2~ sink of temperate deciduous forests.
Rather, our findings imply that *the extra CO~2~ sequestered in years with warmer springs[@keeling_increased_1996;@dragoni_evidence_2011;@crabbe_extreme_2016] is not allocated to the long-lived woody biomass C pool and is therefore unlikely to have a lasting impact on the forest C balance.*
Thus, warming spring temperatures are unlikely to increase the woody productivity or strengthen the carbon sink of temperate deciduous forests.

@@

**Growth milestones were achieved earlier at SCBI than at Harvard Forest (average difference: 6-10 days) and differed between ring- and diffuse-porous species (`r  EDF_growth_curves_all`, `r  EDT_mean_growth_parameters`).**
Across all trees and years, the mean $DOY_{25}$ occurred on day 139 (May 20) at SCBI and day 148 (May 29) at Harvard Forest, while $DOY_{75}$ occurred on day 185 (July 5) at SCBI and day 193 (July 13) at Harvard Forest.
Consistent with the results of Ref[D'Orangeville], ring-porous species began growing earlier, reaching the $DOY_{25}$ benchmark earlier (by `r 154-123` days at SCBI and `r 164-132` at Harvard Forest), and their growth was spread out over a longer growing season (average $L_{PGS}$ 21 and 19 days longer at SCBI and Harvard Forest, respectively). 
Growing season length was similar across sites, with $L_{PGS}$ being, on average, only two days longer at SCBI for ring-porous species and less than one day longer for diffuse-porous species.

@@

For the majority of weekly time windows from Jan. 1 to $DOY_{25}$, there was a negative or neutral relationship between $T_{max}$ and $DOY_{25}$, meaning that warmer temperatures shifted growth earlier in the season (`r EDF_climwin`). 
This held true for both ring- and diffuse-porous trees at both sites, *albiet* with less consistent patterns at Harvard Forest, which was sampled for fewer years. 
The strongest relationships, including but not limited to the optimal time windows (`r  EDT_mean_growth_parameters`), occurred for time windows closing $\le$ 5 weeks prior to $DOY_{25}$.

@@

**To understand how annual growth, as registered in ring width indices ($RWI$), have responded to spring temperatures at the centennial scale, we analyzed tree-ring chronologies for `r n_chronologies_SCBI` species at SCBI, `r n_chronologies_HF` species at Harvard Forest.**
*(The following should be condensed:)*
Residual chronologies rarely exhibited significant correlations to either $T_{max}$ or $T_{min}$ in the months corresponding to the critical temperature windows identified here (Fig. `r tree_rings`, `r EDF_tree_rings_TMIN`).
Although correlations were rarely significant, there was a consistent pattern in responses to $T_{max}$ across all species at both sites:
responses to $T_{max}$ were negative or neutral in all months from February to August, with the exception of one spring month at each site (April at SCBI, May at Harvard Forest) where growth displayed a positive or neutral responses to $T_{max}$ (Fig. `r tree_rings`).
In these months, positive correlations to $T_{max}$ were statistically significant (at p=0.05) for one of nine ring-porous species at SCBI and one of two diffuse-porous species at each site.
In the prior months, the generally negative correlations to $T_{max}$ were not significant, excepting significant correlations of one species at each site  (ring-porous *Carya tomentosa* in March at SCBI; diffuse-porous *Acer rubrum* in April at Harvard Forest).
Starting in May at SCBI, significant correlations to $T_{max}$ were common, occurring for nine of the eleven species in May and 3-7 species for each month June through August.
At Harvard Forest, there was only one statistically significant negative correlation to post-May $T_{max}$ (*Quercus rubra* in June).

@@

Patterns in $T_{min}$ were fairly similar to those of $T_{max}$, but weaker and more stochastic, with more (non-significant) correlations going in opposite directions (`r EDF_tree_rings_TMIN`). 
There were no significant correlations to $T_{min}$ prior to May at SCBI or June at Harvard Forest.

@@

Metal dendrometer bands installed on a total of `r n_dendro_trees` trees within the SCBI and Harvard Forest ForestGEO plots (`r  EDT_species`). 
Bands were placed on dominant species, including two diffuse- and two ring-porous species at SCBI and eight diffuse- and three ring-porous species at Harvard Forest.
Bands were measured with a digital caliper approximately every 1-2 weeks within the growing season from 2011-2020 at SCBI and 1998-2003 at Harvard Forest, respectively. 
The number of bands measured at each site fluctuated somewhat as trees were added or dropped (e.g., because of tree mortality) from the census. 
Across years, the number of bands sampled averaged *~135* (range: 99-145) at SCBI and  *~717* (range: *700-755*) at Harvard Forest. 
In total, our analysis included `r n_tree_years` tree-years (`r  EDT_species`).

@@

### Tree-rings
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (`r  EDT_species`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI, cores were collected as described in @helcoski_growing_2019. 
Briefly, cores were taken from living or recently dead trees $\ge$ 10 cm $DBH$, including a randomly sampled subset of all dominant species within the ForestGEO plot [live trees in 2010-2011; @bourg_initial_2013] and trees found newly dead in annual tree mortality in 2016 and 2017 [@helcoski_growing_2019].
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed `r n_cores_focal_sites` cores from eleven species at SCBI and three species at Harvard Forest (`r  EDT_species`).

@@

All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
We standardized ring-width series from each core via ARSTAN using a 2/3rds $n$ spline, where $n$ is the number of years in the series [@cook_time_1985; @cook_methods_1990]. 
*(The following italic text is self-plagarized from Helcoski and needs to be reworded:)* *The influence of outliers in all series was reduced using the adaptive power transformation, which also stabilises the variance over time [@cook_calculating_1997]. Next, each series was stabilised using either the average correlation between raw ring-width series (rbar) method or a 1/3rds spline method to adjust changes in variance as series replication decreased towards the earlier portion of each chronology [@jones_estimating_1997]. The 1/3rds spline method was chosen when replication in the inner portion of each chronology (c. the inner 30–50 yr of each record depending on full chronology length) dropped below three trees. Once that step was complete, a robust biweight mean chronology for each species was calculated from the ring-width indices [@cook_time_1985]. We chose to use residual chronologies because the autoregressive standardisation process in creating them removes much of the tree-level autocorrelation in growth and these chronologies would most likely contain the most conservative information on drivers of interannual growth [@cook_time_1985].* 
We defined chronology start year (`r  EDT_species`) as the year where subsample signal strength (SSS) passed a threshold of SSS = 0.8, or where $\ge$ 80% of the population signal was captured in the chronology.

@@

The period where the phenology milestone(s) were most affected by changes in early spring temperature was determined using the R package *climwin* [@vandepol_identifying_2016]. 
This package tests the correlation between one or more climate variables and a biological variable over all consecutive time windows within a specified time-frame, reporting correlation and \$Delta AIC$ of each window. 
Here, we tested for correlation between temperature ($T_{max}$, $T_{min}$) and $DOY_{25}$ over the time window from January 1 to the mean $DOY_{25}$ for the species group and site (`r  EDT_mean_growth_parameters`).
*The critical time window is determined based on the subset of models that fall within the 95% confidence set.*
Here, we focused on defining critical temperature windows for $DOY_{25}$, as opposed to other growth phenology parameters (Figure `r schematic_summary`), because we were interested in the influence of temperature on the initiation of woody growth. 
We note that other growth parameters would be controlled by different critical temperature windows; for example, at SCBI $\Delta DBH$ responds most strongly to May-July climate [@helcoski_growing_2019]. Determining the climate variables that most strongly control all parameters is beyond the scope of this analysis.
